Semantic authoring in a flexiformal context --- Bulk annotation of rigorous documents

### 2025-05-13 doc/cicm25-paper/paper.tex
more

\begin{newpart}{MK: please re-read}
  A distinctive feature of semantic authoring that is not apparent at the first glance is
  that the added investment in annotation moves the sweet spot in the re-use/customize
  trade-off towards sharing and interoperability. In our experience sufficiently annotated
  documents can be automatically customized, if the semantics has multiple
  notations/verbalizations/presentations. Consider for instance course materials on
  complex arithmetic, which become re-usable across the electrical
  engineering/mathematics border, if the imaginary unit in formulae is represented e.g. by
  a macro \lstinline|\imaginaryunit|, which can alternatively expand to $i$ (for uses in
  maths) and $j$ (for EE uses). As a consequence semantic authoring is more geared towards
  creating reusable/shareable document fragments organized in central or federated
  repositories and hubs. This is quite similar to the formal authoring system, where we
  see a tendency towards massively shared math libraries in formal systems and package
  hubs for programming libraries. As a consequence, semantic authoring support systems
  should facilitate the logistics of using and curating shared libraries. 
\end{newpart}

\paragraph{Semantic Authoring and the Semantic Web}
The phrase ``semantic authoring'' has also been used in the context of the semantic web with varying meanings.
This includes ontology authoring, which we would classify as formal authoring,
and metadata annotation (e.g. the title, author, and creation date of a document),
which is a somewhat different concern.
A closer match is the annotation of informal content with
references to a domain ontology,
but there are some differences to our situation with \sTeX/\ALeA:
The domain ontology would be a fully formal target
-- possibly containing verbalizations --
while in our case the target is flexiformal.
Furthermore, \sTeX has a module system, which requires
dependency management during authoring.

@@

\begin{newpart}{MK: rewritten as existing but experimental. This needs to be placed
    somewhere else.}
  \paragraph{Scaling \snify to the Flexiformal Content Commons }
One practical problem with \snify system as presented so far is that it only harvests
symbol/verbalization pairs from \sTeX files on the local file system. This means that, for
effective annotation, users (have to) download the whole content commons and keep it
updated -- an assumption/requirement that becomes increasingly impractical as the \sTeX
ecosystem grows.

@@

Note that implemented this broadly \snify can provide additional logistic services: If a
pair is chosen in an annotation, \snify can then automatically download the corresponding
archive so that the annotated file can be formatted with \texttt{pdflatex} or
\textsf{rus\TeX} locally.
\end{newpart}

@@

For less complete domain models we have to skip too many terms that should ultimately be
annotated and annotation efficiency suffers.\ednote{
    FS: Isn't this wrong? We don't have to skip them, the problem is that \snify skips them.
    Efficiency does not suffer, but rather effectiveness.
}
This is currently the case for all
non-English languages in the \sTeX corpus we work with. Coverage in German is only about
half of that of English and we can already see the practical effects.
The \defianno tool (see end of \cref{sec:snify}) can mitigate this partially by
providing verbalizations for the definienda in the document,
but we also want to annotate terms that are not explicitly defined in the document.
A potential solution would be to extend \snify with a machine learning-based
tool for identifying unannotated technical terms that are not in the catalog.
The user could then pick a symbol for annotation, which would add a verbalization to the catalog.
In an earlier (pre-\snify) attempt, we used named-entity
recognition (NER) for classification of ``likely annotation candidate words''
\cite{hutterer:msc23}. However, this classification was not precise enough in the
distinction in ``technical terms'' and ordinary English noun phrases
to be practical on its own.
A combination with \snify's catalog-based approach might change the trade-offs involved.

>>>>

\begin{newpart}{MK: please re-read}
  A distinctive feature of semantic authoring that is not apparent at the first glance is
  that the added investment in annotation moves the sweet spot in the re-use/customize
  trade-off towards sharing and interoperability. In our experience sufficiently annotated
  documents can be automatically customized, if the semantics has multiple
  notations/verbalizations/presentations. Consider for instance course materials on
  complex arithmetic, which become re-usable across the electrical engineering/mathematics
  border, if the imaginary unit in formulae is represented e.g. by a semantic macro
  \lstinline|\imaginaryunit|, which can alternatively expand to $i$ (for uses in maths)
  and $j$ (for EE uses). As a consequence semantic authoring is more geared towards
  creating reusable/shareable document fragments organized in central or federated
  repositories and hubs. This is quite similar to the formal authoring system, where we
  see a tendency towards massively shared math libraries in formal systems and package
  hubs for programming libraries. As a consequence, semantic authoring support systems
  should facilitate the logistics of using and curating shared libraries.
\end{newpart}

\paragraph{Semantic Authoring and the Semantic Web}
The phrase ``semantic authoring'' has also been used in the context of the semantic web
with varying meanings.  This includes ontology authoring, which we would classify as
formal authoring, and metadata annotation (e.g. the title, author, and creation date of a
document), which is a somewhat different concern.  A closer match is the annotation of
informal content with references to a domain ontology, but there are some differences to
our situation with \sTeX/\ALeA: The domain ontology would be a fully formal target --
possibly containing verbalizations -- while in our case the target is flexiformal.
Furthermore, \sTeX has a module system, which requires dependency management during
authoring.\ednote{MK: We need more citations here}

@@

  \paragraph{Scaling \snify to the Flexiformal Content Commons }
One practical problem with \snify system as presented so far is that it only harvests
symbol/verbalization pairs from \sTeX files on the local file system. This means that, for
effective annotation, users (have to) download the whole content commons and keep it
updated -- an assumption/requirement that becomes increasingly impractical as the \sTeX
ecosystem grows.

@@

Note that implemented this broadly \snify can provide additional logistic services: If a
pair is chosen in an annotation, \snify can then automatically download the corresponding
archive so that the annotated file can be formatted with \texttt{pdflatex} or
\textsf{rus\TeX} locally.

@@

For less complete domain models we have to skip too many terms that should ultimately be
annotated and annotation effectiveness suffers. This is currently the case for all
non-English languages in the \sTeX corpus we work with. Coverage in German is only about
half of that of English and we can already see the practical effects.  The \defianno tool
(see end of \cref{sec:snify}) can mitigate this partially by providing verbalizations for
the definienda in the document, but we also want to annotate terms that are not explicitly
defined in the document.  A potential solution would be to extend \snify with a machine
learning-based tool for identifying unannotated technical terms that are not in the
catalog.  The user could then pick a symbol for annotation, which would add a
verbalization to the catalog.  In an earlier (pre-\snify) attempt, we used named-entity
recognition (NER) for classification of ``likely annotation candidate words''
\cite{hutterer:msc23}. However, this classification was not precise enough in the
distinction in ``technical terms'' and ordinary English noun phrases to be practical on
its own.  A combination with \snify's catalog-based approach might change the trade-offs
involved.
